Psychophysics of musical elements in the discrete-time representation of sound

### 2013-10-07 article/article.tex
feito ateh idioma musical

The philosophical understanding that the human thought is based on the perception of
similarities, differences, stimuli and objects, places the symmetries
at the core of the cognitive process~\cite{Deleuze}. Mathematically, symmetries
are algebraic finite groups, and a finite group is always isomorphic to a permutation
group. It is possible to say that permutations represent any symmetry in a
finite system (precisa de uma referência aqui). In music, permutations are ubiquitous and considerably present in techniques, which confirms their central role. The successive application of permutations generates cyclic arcs~\cite{change,Zamacois,permMusic}. Two academic studies were dedicated to this end, aiming at generating music structures~\cite{figgusOriginal, figgusEspacializacao}.

@@

\begin{multline}\label{eq:groups}
%\begin{split}
\forall \;\; p_1,p_2 \in G \Rightarrow\quad\;\; p_1 \bullet p_2  =
p_3 \in G \\ \;\;\;\text{(closure property)} \\
\forall \;\; p_1,p_2,p_3 \in G \Rightarrow\quad (p_1\bullet p_2)\bullet p_3  =
p_1\bullet (p_2\bullet p_3)\quad\;  \\ \text{(associativity property)} \\
\exists \;\; e \in G :\; p \bullet e  = e \bullet
p \;\;\;\; \forall p \in G  \quad \\ \text{(identity element property)} \\
\forall \;\; p \in G, \;\exists\; p^{-1} :\quad\;  p\bullet p^{-1} 
=p^{-1}\bullet p = e \\ \; \text{(inverse element property)}
%\end{split}
\end{multline}

From the first property it is possible to conclude that any permutation can be
operated with another permutation. In fact, it is possible to apply a
permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final
orderings, results in another permutation $p_3$.

Every element $p$ operated with itself a sufficient number of times $n$ reaches
the identity element $p^n=e$ (otherwise the group would be infinite, generated
by $p$). The lower $n\;:\;p^n=e$ is called the element order. Thus, a finite
permutation $p$, successively applied, reaches the initial ordering of its
elements, making a cycle. This cycle, if used as parameter of music notes,
implies in a cyclic musical arc.

These arcs can be established by using a set of different permutations. As a historic
example, the tradition called \emph{change ringing} conceives the music through
bells played one after other and then played again, but in a different
order. This process repeats until it reaches the initial ordering. The set of
different traversed orderings is a \emph{peal}. Table~\ref{tab:change}
represents one traditional \emph{peal} for 3 bells (1, 2 and 3) which explores
all possible orderings. Each line indicates one bell ordering to be
played. Permutations occur between each line. In this case, the music structure
consists of permutations by itself and some different permutations operate due to
the cyclic behavior.

\begin{table}[htpq!]
\centering
\caption{Change Ringing: Standard \emph{peal} for 3 bells. Permutations
occur between each line. Each line is a bell ordering and each ordering is played
        a line a time.} 
\begin{tabular}{l c r}
\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3} \\
\textcolor{blue}{2} & \textcolor{red}{1} & \textcolor{green}{3} \\
\textcolor{blue}{2} & \textcolor{green}{3} & \textcolor{red}{1} \\
\textcolor{green}{3} & \textcolor{blue}{2} & \textcolor{red}{1} \\
\textcolor{green}{3} & \textcolor{red}{1} & \textcolor{blue}{2} \\
\textcolor{red}{1} & \textcolor{green}{3} & \textcolor{blue}{2} \\
\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3}
\end{tabular}
\label{tab:change}
\end{table}

The use of permutations in music can be summarized in the following way:
consider $S_i=\{s_i\}$ as a sequence of musical events $s_i$ (e.g.\ notes), and a
permutation $p$. $S_i'=p(S_i)$ comprises the same elements of $S_i$ but in a
different order. The permutations can be written based on two notations: cyclic or
natural. The natural notation basically comprehends the order of the resulting indexes from
the permutation. Thus, agreed the original ordering given the sequence of its
indexes $[0\;1\;2\;3\;4\;5\;...]$, the permutation is noted by the sequence it
produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, permutation is the
changing of one element by the posterior one, and the last element by the first one.

It is not necessary to permute elements of $S_i$, but only some
characteristics. In this way, being $p^f$ a permutation in frequency and $S_i$ a
sequence of basic notes as exposed in the end of section~\ref{notaBasica}, the
new sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
music notes, respecting same order and maintaining the same characteristics, but with the
fundamental frequencies permuted following the pattern that $p^f$ presents.

It is worthwhile to mention two subtleties of this procedure. First, the permutation $p$ does not have to involve all elements of $S_i$, i.e.\ it can operate in subsets of $S_i$. Second,  the elements $s_i$ do not have to be entirely executed at each state access. To exemplify,
consider $S_i$ as a sequence of music notes $s_i$. If $i$ goes from $0$ to $n$, and
$n>4$, at each measure of $4$ notes it is possible to execute the first $4$
notes. The other notes of $S_i$ can occur in measures where permutations set
aside those notes to the first four notes of $S_i$.

Each one of the permutations $p_i$ described above relates the note dimensions where it operates (frequency, duration, \emph{fades},
intensity, etc) and the period of incidence (at how much access a permutation is
applied). During realization of notes in $S_i$, an easy and coherent form is to
execute the first $n$ notes\footnote{The execution of disjoint notes of $S_i$
equals to modify the permutation and to execute the first notes.}.

Appendix~\ref{cap:FIGGUScode} and \massa present the computational implementation about permutations~\cite{MASSA,figgusOriginal,figgusEspacializacao}.

@@

>>>>

The philosophical understanding that human thought is founded on the concept of
similarities and differences (e.g. perceived in stimuli and objects), places symmetries
at the core of cognitive processes~\cite{Deleuze}. Mathematically, symmetries
are algebraic groups, and a finite group is always isomorphic to a permutation
group (by Cayley's theorem). 
In a way, this states that permutations represent any symmetry in a
finite system~\cite{grupos,gruposFascination}.
In music, permutations are ubiquitous and present in scholastic techniques,
 which confirms their central role. 
The successive application of permutations generates cyclic arcs~\cite{change,Zamacois,permMusic}. 
Two academic studies were dedicated to generating musical structures~\cite{figgusOriginal, figgusEspacializacao}.

@@

\begin{equation}\label{eq:groups}
\begin{split}
\forall \;\; p_1,p_2 \in G \Rightarrow  \quad   & p_1 \bullet p_2  = p_3 \in G \\ 
     & \text{(closure property)} \\
\forall \;\; p_1,p_2,p_3 \in G \Rightarrow \quad & (p_1\bullet p_2)\bullet p_3  = p_1\bullet (p_2\bullet p_3) \\
     & \text{(associativity property)} \\
\exists \;\; e \in G :                  \quad    & p \bullet e  = e \bullet p \;\;\;\; \forall p \in G  \\ 
     &  \text{(identity element existence)} \\
\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1}i =p^{-1}\bullet p = e \\
     &  \text{(inverse element existence)}
\end{split}
\end{equation}

From the first property follows that two permutations act as one permutation.
In fact, it is possible to apply a
permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final
orderings, observe another permutation $p_3$.

Every element $p$ operated with itself a sufficient number of times $n$ reaches
the identity element $p^n=e$ (otherwise the group would be infinite, generated
by $p$). The lower $n\;:\;p^n=e$ is the element order. Thus, a finite
permutation $p$, successively applied, reaches the initial ordering of its
elements, and makes a cycle. This cycle, if used for parameters of musical notes or structures,
yields a cyclic musical arc.

These arcs can be established by using a set of different permutations. As a historic
example, the \emph{change ringing} tradition conceives music through
bells played one after other and then played again, but in a different
order. This process repeats until it reaches the initial ordering. The set of
different traversed orderings is a \emph{peal}. Table~\ref{tab:change}
presents a traditional \emph{peal}, named ``Hunt'', for 3 bells (\textcolor{red}{1}, \textcolor{blue}{2} and \textcolor{green}{3}), which explores
all possible orderings. Each line indicates one bell ordering to be
played. Permutations occur between each line. In this case, music structure
consists of permutations itself and some different permutations operate in the
cyclic behavior.

\begin{table}[htpq!]
\centering
\caption{Change Ringing: Traditional \emph{peal} for 3 bells. Permutations
occur between each line. Each line is a bell ordering and each ordering is played at a time.} 
\begin{tabular}{l c r}
\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3} \\
\textcolor{blue}{2} & \textcolor{red}{1} & \textcolor{green}{3} \\
\textcolor{blue}{2} & \textcolor{green}{3} & \textcolor{red}{1} \\
\textcolor{green}{3} & \textcolor{blue}{2} & \textcolor{red}{1} \\
\textcolor{green}{3} & \textcolor{red}{1} & \textcolor{blue}{2} \\
\textcolor{red}{1} & \textcolor{green}{3} & \textcolor{blue}{2} \\
\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3}
\end{tabular}
\label{tab:change}
\end{table}

The use of permutations in music can be summarized in the following way:
with $S_i=\{s_i\}$ a sequence of musical events $s_i$ (e.g.\ notes), and a
permutation $p$. $S_i'=p(S_i)$ comprises the same elements of $S_i$ but in a
different order. Permutations has two notations: cyclic and
natural. The natural notation basically comprehends the order of the resulting indexes from
the permutation. Thus, agreed the original ordering given the sequence of its
indexes $[0\;1\;2\;3\;4\;5\;...]$, the permutation is noted by the sequence it
produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, a permutation consists
of swaps by elements and its posteriors, and the last element by the first one.
E.g. $(1,2,5)(3,4)$ in cyclic notation is equivalent to $[0,2,5,4,3,1]$ in natural notation.

In the auralization of a permutation, it is not necessary to permute elements of $S_i$,
but only some characteristic. In this way, being $p$ a permutation and $S_i$ a
sequence of basic notes as exposed in the end of section~\ref{notaBasica}, the
sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
music notes, respecting same order and maintaining the same characteristics, but with the
fundamental frequencies permuted following the pattern that $p$ presents.

Two subtleties of this procedure. 
First, a permutation $p$ is not restricted to involve all elements of $S_i$,
i.e.\ it can operate in a subset of $S_i$.
Second,  not all elements $s_i$ need to be executed at each access to $S_i$.
To exemplify, be $S_i$ a sequence of music notes $s_i$. 
If $i$ goes from $0$ to $n$, and
$n>4$, at each measure of $4$ notes it is possible to execute the first $4$
notes. The other notes of $S_i$ can occur in measures where permutations 
allocate those notes to the first four events $s'_i$ of $S'_i$.

It is inalienable to each permutations $p_i$ described above:
1) note dimensions where it operates (frequency, duration, \emph{fades},
intensity, etc); and
2) the period of incidence (how much consults before a permutation is
applied). During realization of notes in $S_i$, an easy and coherent form is to
execute the first $n$ notes, the execution of disjoint sets of $S_i$
equals modifying the permutation and executing the first $n$ notes.

The \massa\ toolbox presents a computational implementation that isolates
permutation aplication to sonic characteristics, in order to deliver
musical structures~\cite{MASSA,figgusOriginal,figgusEspacializacao}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
